---
title: "Homework 3"
author: "Mukta Patwari"
date: "2025-10-09"
output: github_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(p8105.datasets)

knitr::opts_chunk$set(
  fig.width = 6,
  fig.asp = .6,
  out.width = "90%"
)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```

## Problem 1

Loading and inspecting data
```{r}
data("instacart")
```

There are `r nrow(instacart)` observations in this dataset, with the following `r ncol(instacart)` variables: `r colnames(instacart)`. Some of the key variables are `order_id` and `user_id` which give information on the individual ordering these items on Instacart. Other key variables are `product_name`, which refers to what product(s) the individual ordered, and `aisle` and `department` which give location data within the grocery store.

**How many aisles are there, and which aisles are the most items ordered from?**

```{r}
instacart %>% 
  distinct(aisle) %>% 
  nrow()

instacart %>% 
  group_by(aisle) %>% 
  summarize(
    n = n()
  ) %>% 
  arrange(desc(n))
```

There are 134 different aisles in this dataset. The top 10 most frequented aisles were: fresh vegetables, fresh fruits, packaged vegetables/fruits, yogurt, packaged cheese, water/seltzer/sparkling water, milk, chips/pretzels, soy/lactose free, and bread.

**Make a plot that shows the number of items ordered in each aisle, limiting this to aisles with more than 10000 items ordered. Arrange aisles sensibly, and organize your plot so others can read it.**

```{r aisle-plot}
instacart %>% 
  group_by(aisle) %>% 
  summarize(
    n = n()
  ) %>%
  filter(n > 10000) %>%
  ggplot(aes(x = reorder(aisle, n), y = n)) +
  geom_point() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) +
  labs(
    title = "Items Ordered by Aisle (>10,000 Orders)",
    x = "Aisle Name",
    y = "Number of Items Ordered"
  )
```

This plot corroborates the results that were seen in the earlier question, about how many items were ordered from each aisle which aisles had the most items ordered - fresh vegetables has a high item count at greater >120000 orders, as does the fresh fruits category.

**Make a table showing the three most popular items in each of the aisles “baking ingredients”, “dog food care”, and “packaged vegetables fruits”. Include the number of times each item is ordered in your table.**

```{r}
instacart %>%
  filter(aisle %in% c("baking ingredients", "dog food care", "packaged vegetables fruits")) %>% 
  count(aisle, product_name) %>%
  group_by(aisle) %>% 
  mutate(product_rank = min_rank(desc(n))) %>% 
  filter(product_rank <= 3) %>%
  arrange(aisle, product_rank) %>%
  select(-product_rank) %>% 
  knitr::kable(digits = 2)
```

The table shows that in the baking ingredidents aisle, the products light brown sugar, pure baking soda, and cane sugar were most ordered; in the dog food care aisle, the most ordered products were "Snack Sticks Chicken & Rice Recipe Dog Treats", "Organix Chicken & Brown Rice Recipe" and small dog biscuits; in the packaged vegetables fruits aisle, the most ordered items were organic baby spinach, organic raspberries, and organic blueberries.

**Make a table showing the mean hour of the day at which Pink Lady Apples and Coffee Ice Cream are ordered on each day of the week; format this table for human readers (i.e. produce a 2 x 7 table).**

```{r}
instacart %>%
  mutate(order_dow = factor(order_dow, 
    levels = 0:6,
    labels = c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))) %>% 
  filter(product_name %in% c("Pink Lady Apples", "Coffee Ice Cream")) %>%
  group_by(product_name, order_dow) %>%
  summarize(mean_hour = mean(order_hour_of_day, na.rm = TRUE)) %>% 
  pivot_wider(
    names_from = order_dow,
    values_from = mean_hour
  ) %>% 
  knitr::kable(digits = 1)
```

This table shows the mean order hour of day for each day of the week that the coffee ice cream and pink lady apples were ordered. On Sunday, coffee ice cream was ordered at hour 13.8, which when converted to an hour-time format, would probably be around 1:40. Similarly, the mean hour that pink lady apples were ordered was 13.4, which would be around 1:20. The remaining columns are the days of the week Monday-Saturday, with the corresponding mean hours for the same products ordered.

## Problem 2

Loading, importing, tidying zipcodes
```{r}
zipcodes_df =
  read_csv(
    "zillow_data/Zip Codes.csv", 
    na = c("NA", ".", "")
    ) %>%
  janitor::clean_names() %>%
  filter(!(zip_code %in% c(10463, 11201) & county == "New York")) %>% 
  select(-state_fips, -county_code, -county_fips, -file_date)


# Duplicated zip-codes:
zipcodes_df[duplicated(zipcodes_df$zip_code),]
```

Loading, importing, tidying rental price
```{r}
rentalprice_df =
  read_csv("zillow_data/Zip_zori_uc_sfrcondomfr_sm_month_NYC.csv", na = c("NA", ".", "")) %>% 
  rename_with(janitor::make_clean_names, .cols = 1:9) %>% 
  pivot_longer(
    cols = "2015-01-31":"2024-08-31",
    names_to = "date",
    values_to = "rent",
  ) %>%
  mutate(
    date = as.Date(date)
  ) %>%
  separate(
    date, into = c("year", "month", "day"),
    convert = TRUE,
  ) %>% 
  rename(zip_code = region_name, county = county_name) %>%
  mutate(
    county = (str_remove(county, "County"))
  ) %>% 
  select(-region_id, -size_rank, -region_type, -state_name, -metro, -county)
```

Merging datasets into zillow dataframe
```{r}
zillow_df =
  left_join(
    rentalprice_df, zipcodes_df, by = "zip_code"
    ) %>%
  relocate(year, month, day, state, city, county, zip_code, neighborhood, rent)
```

**How many ZIP codes are observed 116 times? How many are observed fewer than 10 times? Why are some ZIP codes are observed rarely and others observed in each month?**

```{r}
zillow_df %>%
  drop_na() %>% 
  group_by(zip_code) %>% 
  summarise(count = n()) %>% 
  summarise(
    sum(count == 116),
    sum(count < 10)
  )
```

There are 47 zip codes that appear 116 times throughout the whole dataset, and 25 zip codes that appear less than 25 times. This is likely related to population densities and urbanization. The zip code 10001 is distinctly in Manhattan, which is a highly urban area with many people and large fluctuations in rental prices/unit occupancies. It is very common for people to sign one-year leases and then not renew their licenses, which means these units are put on the market every year. On the other hand, a zip code like 10308, which is in Staten Island may show up only a few times as these kinds of areas are more suburban and resiential, and occupants may rent these places for years together (therefore no rental prices every year).

**Create a reader-friendly table showing the average rental price in each borough and year.**

```{r}
zillow_df %>%
  group_by(county, year) %>%
  summarize(
    mean_rent = mean(rent, na.rm = TRUE)
  ) %>% 
  pivot_wider(
    names_from = "county",
    values_from = "mean_rent"
  ) %>% 
  knitr::kable(digits = 2)
```

Since 2015, there has been a general increase in rent. While the rental price observations are unavailable in Richmond county from 2015-2019, there is still a clear increase, especially after 2020. The highest rental price increases occur in New York county/Manhattan, from 3022.04 dollars in 2015 to 4078.44 dollars in 2024. A more noticeable trend is how rental prices changed during/after the pandemic, and that they increased immensely. This is probably because people were not living in the city due to lockdown, and to compensate, realty groups had to increase their prices significantly due to losses. 

**Make a plot showing NYC Rental Prices within ZIP codes for all available years.**

```{r zip-codes-plot}
zillow_df %>%
  group_by(zip_code, county, year) %>% 
  summarize(
    mean_rent = mean(rent, na.rm = TRUE)
  ) %>% 
  ggplot(aes(x = year, y = mean_rent, color = county)) +
  geom_point() + 
  facet_grid(. ~ county) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) +
  labs(
    title = "NYC Rental Prices Across Borough, from 2015-2024",
    x = "Year",
    y = "Mean Rental Price"
  )
```