---
title: "Homework 3"
author: "Mukta Patwari"
date: "2025-10-09"
output: github_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(p8105.datasets)

knitr::opts_chunk$set(
  fig.width = 6,
  fig.asp = .6,
  out.width = "90%"
)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d

library(patchwork)
```

## Problem 1

Loading and inspecting data
```{r}
data("instacart")
```

There are `r nrow(instacart)` observations in this dataset, with the following `r ncol(instacart)` variables: `r colnames(instacart)`. Some of the key variables are `order_id` and `user_id` which give information on the individual ordering these items on Instacart. Other key variables are `product_name`, which refers to what product(s) the individual ordered, and `aisle` and `department` which give location data within the grocery store.

**How many aisles are there, and which aisles are the most items ordered from?**

```{r}
instacart %>% 
  distinct(aisle) %>% 
  nrow()

instacart %>% 
  group_by(aisle) %>% 
  summarize(
    n = n()
  ) %>% 
  arrange(desc(n))
```

There are 134 different aisles in this dataset. The top 10 most frequented aisles were: fresh vegetables, fresh fruits, packaged vegetables/fruits, yogurt, packaged cheese, water/seltzer/sparkling water, milk, chips/pretzels, soy/lactose free, and bread.

**Make a plot that shows the number of items ordered in each aisle, limiting this to aisles with more than 10000 items ordered. Arrange aisles sensibly, and organize your plot so others can read it.**

```{r aisle-plot}
instacart %>% 
  group_by(aisle) %>% 
  summarize(
    n = n()
  ) %>%
  filter(n > 10000) %>%
  ggplot(aes(x = reorder(aisle, n), y = n)) +
  geom_point() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) +
  labs(
    title = "Items Ordered by Aisle (>10,000 Orders)",
    x = "Aisle Name",
    y = "Number of Items Ordered"
  )
```

This plot corroborates the results that were seen in the earlier question, about how many items were ordered from each aisle which aisles had the most items ordered - fresh vegetables has a high item count at greater >120000 orders, as does the fresh fruits category.

**Make a table showing the three most popular items in each of the aisles “baking ingredients”, “dog food care”, and “packaged vegetables fruits”. Include the number of times each item is ordered in your table.**

```{r}
instacart %>%
  filter(aisle %in% c("baking ingredients", "dog food care", "packaged vegetables fruits")) %>% 
  count(aisle, product_name) %>%
  group_by(aisle) %>% 
  mutate(product_rank = min_rank(desc(n))) %>% 
  filter(product_rank <= 3) %>%
  arrange(aisle, product_rank) %>%
  select(-product_rank) %>% 
  knitr::kable(digits = 2)
```

The table shows that in the baking ingredidents aisle, the products light brown sugar, pure baking soda, and cane sugar were most ordered; in the dog food care aisle, the most ordered products were "Snack Sticks Chicken & Rice Recipe Dog Treats", "Organix Chicken & Brown Rice Recipe" and small dog biscuits; in the packaged vegetables fruits aisle, the most ordered items were organic baby spinach, organic raspberries, and organic blueberries.

**Make a table showing the mean hour of the day at which Pink Lady Apples and Coffee Ice Cream are ordered on each day of the week; format this table for human readers (i.e. produce a 2 x 7 table).**

```{r}
instacart %>%
  mutate(order_dow = factor(order_dow, 
    levels = 0:6,
    labels = c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))) %>% 
  filter(product_name %in% c("Pink Lady Apples", "Coffee Ice Cream")) %>%
  group_by(product_name, order_dow) %>%
  summarize(mean_hour = mean(order_hour_of_day, na.rm = TRUE)) %>% 
  pivot_wider(
    names_from = order_dow,
    values_from = mean_hour
  ) %>% 
  knitr::kable(digits = 1)
```

This table shows the mean order hour of day for each day of the week that the coffee ice cream and pink lady apples were ordered. On Sunday, coffee ice cream was ordered at hour 13.8, which when converted to an hour-time format, would probably be around 1:40. Similarly, the mean hour that pink lady apples were ordered was 13.4, which would be around 1:20. The remaining columns are the days of the week Monday-Saturday, with the corresponding mean hours for the same products ordered.

## Problem 2

Loading, importing, tidying zipcodes
```{r}
zipcodes_df =
  read_csv(
    "zillow_data/Zip Codes.csv", 
    na = c("NA", ".", "")
    ) %>%
  janitor::clean_names() %>%
  filter(!(zip_code %in% c(10463, 11201) & county == "New York")) %>%
  mutate(
    borough = case_match(
      county,
      "Bronx" ~ "Bronx",
      "Kings" ~ "Brooklyn",
      "New York" ~ "Manhattan",
      "Queens" ~ "Queens",
      "Richmond" ~ "Staten Island"
    )) %>% 
  select(-state_fips, -county_code, -county_fips, -file_date, -county)


# Duplicated zip-codes:
zipcodes_df[duplicated(zipcodes_df$zip_code),]
```

Loading, importing, tidying rental price
```{r}
rentalprice_df =
  read_csv("zillow_data/Zip_zori_uc_sfrcondomfr_sm_month_NYC.csv", na = c("NA", ".", "")) %>% 
  rename_with(janitor::make_clean_names, .cols = 1:9) %>% 
  pivot_longer(
    cols = "2015-01-31":"2024-08-31",
    names_to = "date",
    values_to = "rent",
  ) %>%
  mutate(
    date = as.Date(date)
  ) %>%
  separate(
    date, into = c("year", "month", "day"),
    convert = TRUE,
  ) %>% 
  rename(zip_code = region_name, county = county_name) %>%
  mutate(
    county = (str_remove(county, "County"))
  ) %>% 
  select(-region_id, -size_rank, -region_type, -state_name, -metro, -county)
```

Merging datasets into zillow dataframe
```{r}
zillow_df =
  left_join(
    rentalprice_df, zipcodes_df, by = "zip_code"
    ) %>%
  relocate(year, month, day, state, city, borough, zip_code, neighborhood, rent)
```

**How many ZIP codes are observed 116 times? How many are observed fewer than 10 times? Why are some ZIP codes are observed rarely and others observed in each month?**

```{r}
zillow_df %>%
  drop_na() %>% 
  group_by(zip_code) %>% 
  summarise(count = n()) %>% 
  summarise(
    sum(count == 116),
    sum(count < 10)
  )
```

There are 47 zip codes that appear 116 times throughout the whole dataset, and 25 zip codes that appear less than 25 times. This is likely related to population densities and urbanization. The zip code 10001 is distinctly in Manhattan, which is a highly urban area with many people and large fluctuations in rental prices/unit occupancies. It is very common for people to sign one-year leases and then not renew their licenses, which means these units are put on the market every year. On the other hand, a zip code like 10308, which is in Staten Island may show up only a few times as these kinds of areas are more suburban and resiential, and occupants may rent these places for years together (therefore no rental prices every year).

**Create a reader-friendly table showing the average rental price in each borough and year.**

```{r}
zillow_df %>%
  group_by(borough, year) %>%
  summarize(
    mean_rent = mean(rent, na.rm = TRUE)
  ) %>% 
  pivot_wider(
    names_from = "borough",
    values_from = "mean_rent"
  ) %>% 
  knitr::kable(digits = 2)
```

Since 2015, there has been a general increase in rent. While the rental price observations are unavailable in Staten Island from 2015-2019, there is still a clear increase, especially after 2020. The highest rental price increases occur in Manhattan, from 3022.04 dollars in 2015 to 4078.44 dollars in 2024. A more noticeable trend is how rental prices changed during/after the pandemic, and that they increased immensely. This is probably because people were not living in the city due to lockdown, and to compensate, realty groups had to increase their prices significantly due to losses. 

**Make a plot showing NYC Rental Prices within ZIP codes for all available years.**

```{r zip-codes-plot}
zipcodes_price_annual = 
  zillow_df %>%
  group_by(zip_code, borough, year) %>%
  summarize(
    mean_rent = mean(rent, na.rm = TRUE)
  ) %>% 
  ggplot(aes(x = year, y = mean_rent, color = borough)) +
  geom_point(alpha = 0.5) +
  facet_grid (. ~ borough) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) +
  labs(
    title = "New York City Rental Prices Across Borough, from 2015-2024",
    x = "Year",
    y = "Average Rental Price"
  ) +
  scale_x_continuous(
    breaks = seq(2015, 2024, by = 3)
  )

print(zipcodes_price_annual)
```

This plot shows the distribution of average rental prices, for each zip-code from 2015-2024, divided by borough. This shows that there is a general upwards trend in zip code-level rental price from 2015-2024 across each borough, with most significant increases in Manhattan, folowed by Queens and then Brooklyn. 

**Compute the average rental price within each ZIP code over each month in 2023. Make a reader-friendly plot showing the distribution of ZIP-code-level rental prices across boroughs.**

```{r 2023-plot}
zipcodes_price_2023 = 
  zillow_df %>%
  group_by(zip_code, borough, month) %>% 
  filter(year == 2023) %>%
  mutate(
    month = month.abb[(month)],
    month = factor(month, levels = month.abb)
  ) %>% 
  summarize(
    mean_rent_2023 = mean(rent, na.rm = TRUE)
  ) %>%
  ggplot(aes(x = month, y = mean_rent_2023, color = borough)) +
  geom_point(alpha = 0.5) +
  facet_grid (. ~ borough) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) +
  labs(
    title = "New York City Rental Prices Across Borough in 2023",
    x = "Month",
    y = "Average Rental Price"
  )

print(zipcodes_price_2023)
```

This plot shows the distribution of average rental prices, for each zip-code from 2023, divided by borough. Here, the trend is relatively stable with no large significant increases or decreases at the zip-code level. Even in the year of 2023, Manhattan had some of the highest average rental prices when compared to the other boroughs.

**Combine the two previous plots into a single graphic, and export this to a results folder in your repository.**

```{r combined-plot}
combined_plots = zipcodes_price_annual/zipcodes_price_2023

ggsave("results/nyc_rent_comparisons.png", combined_plots, width = 12, height = 10, dpi = 300)
```

## Problem 3

Loading, tidying demographic data
```{r}
demo_df =
  read_csv("nhanes_covar.csv", skip = 4) %>% 
  janitor::clean_names() %>% 
  mutate(
    sex = case_match(
      sex,
      1 ~ "male",
      2 ~ "female"
    ),
    education = case_match(
      education,
      1 ~ "Less than high school",
      2 ~ "High school equivalent",
      3 ~ "More than high school"
    ),
    education = factor(education, levels = c("Less than high school", "High school equivalent", "More than high school")),
    sex = factor(sex, levels = c("male", "female"))
  )
```

Loading, tidying accelerometer data
```{r}
accel_df =
  read_csv("nhanes_accel.csv") %>% 
    janitor::clean_names()
```

Merging datasets
```{r}
nhanes_df =
  left_join(demo_df, accel_df, by = "seqn") %>% 
  drop_na(sex, age, bmi, education) %>% 
  filter(age >= 21)
```

**Produce a reader-friendly table for the number of men and women in each education category, and create a visualization of the age distributions for men and women in each education category.**

```{r age-dist-plot}
nhanes_df %>% 
  group_by(sex, education) %>% 
  summarize(
    count = n()
  ) %>%
  pivot_wider(
    names_from = "education",
    values_from = "count"
  ) %>% 
  knitr::kable()

nhanes_df %>% 
  group_by(sex, education) %>% 
  ggplot(aes(x = age, fill = sex)) +
  geom_density(alpha = 0.5) +
  facet_grid(. ~ education) +
  labs(
    title = "Distribution of Age by Sex per Education Category"
  )
```

The table depicts how education levels are different according to sex. Namely, the `more than high school` category had more participants, regardless of sex, although there are just 3 more women comparatively. In the `less than high school` category, the number of men and number of women are almost identical; there are 27 men and 28 women. The "largest" difference is seen in the `high school equivalent`category, where there are 12 more men than women. 

The density plots show the age distributions for these participants, divided by both sex and education category. For those in the `less than high school category`, the densities largely overlap, with most participants in this category ranging between ~40-70 years. There are not as many participants between the ages of 20-40. In the `high school equivalent` category, the distribution for men peaks at around 20-40 years and then again at 60 years, while for women it peaks at around 75-80 years. This indicates that this category has more older women and relatively younger men, and there is less overlap. In the `more than high school` category, the distributions overlap again, with peaks between 20-40 years.Importantly, the density for women is much higher than that of men, indicating that there are more women with higher education in this group than there are men. For this group, there are remarkably less participants at older ages.

**Using your tidied dataset, aggregate across minutes to create a total activity variable for each participant. Plot these total activities (y-axis) against age (x-axis); your plot should compare men to women and have separate panels for each education level. Include a trend line or a smooth to illustrate differences.**

```{r total-activity-plot}
nhanes_df = nhanes_df %>% 
  pivot_longer(
    cols = starts_with("min"),
    names_to = "minutes",
    values_to = "mims_value"
  ) %>% 
  mutate(
  minutes = str_remove(minutes, "min") %>% 
  as.numeric(minutes)
  )

nhanes_df %>% 
  group_by(seqn, age, sex, education) %>% 
  summarize(
    total_activity = sum(mims_value)
  ) %>%
  ggplot(aes(x = age, y = total_activity, color = sex)) +
  geom_point(alpha = 0.5) +
  geom_smooth(se = FALSE) +
  facet_grid(. ~ education) +
  labs(
    title = "Total activity per day by age and educational level",
    x = "Age",
    y = "Total Activity per day"
  )
```

This plot displays age distributions for total activity per day across each educational level, divided by sex. For those in the `less than high school` category, there is an overall decline in total activity as age increases, with slight peaks at age 60. Total activity is higher for women at age 20 than it is for men, and at age 40 this relationship reverses and men consistently have higher total activity numbers. For those in the `high school equivalent` category, the age distribution for women peaks at 40 years and 70 years, while for men it peaks only at 40. Generally, women have higher total activity for all ages when compared to men, except for at 80 years when they level off. For those in the `more than high school` category, there are less obvious peaks and aprubt changes, but there is still a declining trend from 20 years to 80 years. Similar to the `high school equivalent` category, women generally have higher total activity when compared to men. Based on these observations, I would conclude that high school education or more increases total activity in women when compared to men, and that it is highest at 20-40 years before leveling off between 60-80 years.

**Make a three-panel plot that shows the 24-hour activity time courses for each education level and use color to indicate sex.**

```{r 24-hr-activity}
nhanes_df %>% 
  group_by(minutes, sex, education) %>% 
  summarize(
    mean_activity = mean(mims_value)
  ) %>% 
  ggplot(aes(x = minutes, y = mean_activity, color = sex)) +
  geom_point(alpha = 0.1) +
  geom_smooth(aes(group = sex), se = FALSE) +
  facet_grid(. ~ education) +
  scale_x_continuous(
    breaks = seq(0, 1440, by = 360)
  ) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) +
  labs(
    title = "Per day activity-time course by sex and education level",
    x = "Minutes",
    y = "Mean activity"
  )
```

This plot depicts the mean activity per minute against minute to show the per day activity-time course, stratified by educational level and divided by gender.
I chose to look at mean data instead of individual-level data as it makes group-level trends more obvious, and reduces a lot of the noise that resulted from when I initially created the plot with individual data (I went back and revised my plot to include mean activity). Based on this, the mean activities/minute for men and women both majorly overlap for all three educational categories, with women having slightly increased mean activity throughout. When comparing educational levels, those in the `less than high school` category had greater overall mean activity/minute than those in the other two categories. 