---
title: "Homework 3"
author: "Mukta Patwari"
date: "2025-10-09"
output: github_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(p8105.datasets)

knitr::opts_chunk$set(
  fig.width = 6,
  fig.asp = .6,
  out.width = "90%"
)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```

## Problem 1

Loading and inspecting data
```{r}
data("instacart")
```

There are `r nrow(instacart)` observations in this dataset, with the following `r ncol(instacart)` variables: `r colnames(instacart)`. Some of the key variables are `order_id` and `user_id` which give information on the individual ordering these items on Instacart. Other key variables are `product_name`, which refers to what product(s) the individual ordered, and `aisle` and `department` which give location data within the grocery store.

**How many aisles are there, and which aisles are the most items ordered from?**

```{r}
instacart %>% 
  distinct(aisle) %>% 
  nrow()

instacart %>% 
  group_by(aisle) %>% 
  summarize(
    n = n()
  ) %>% 
  arrange(desc(n))
```

There are 134 different aisles in this dataset. The top 10 most frequented aisles were: fresh vegetables, fresh fruits, packaged vegetables/fruits, yogurt, packaged cheese, water/seltzer/sparkling water, milk, chips/pretzels, soy/lactose free, and bread.